---
title: "National Geographies to Elmer"
output: html_notebook
---

This code imports the US Federal Information Processing Standard (FIPS) codes into the central database.
These codes allow for joining to a variety of data sets for spatial display, processing and summarization.

```{r Lookups_to_SQL, include=FALSE}
library(data.table)
library(tidyverse)
library(odbc)
library(DBI)
library(openxlsx)
library(foreign)
library(reticulate)

data_dir <- getwd()
shapefile_dir <- paste0(data_dir,"/shapefiles")

# SQL Database Connection settings
elmer_connection <- dbConnect(odbc::odbc(),
  driver = "SQL Server",
  server = "sql2016\\DSADEV",
  database = "Sandbox",
  trusted_connection = "yes"
  )

tablename <- "fips_codes"

# Create data.tables from the various national coverages that we have shapefiles for
states <- read.dbf(paste0(shapefile_dir,"/cb_2017_us_state_500k.dbf")) %>% as.data.table
counties <- read.dbf(paste0(shapefile_dir,"/cb_2017_us_county_500k.dbf")) %>% as.data.table
cbsa <- read.dbf(paste0(shapefile_dir,"/cb_2017_us_cbsa_500k.dbf")) %>% as.data.table
urban_area <- read.dbf(paste0(shapefile_dir,"/cb_2017_us_ua10_500k.dbf")) %>% as.data.table

# Clean up unique datatables for provide common attribute names
final_column_order <- c('geoid','name','state','category','csa')

# States
states <- states[,c('GEOID','NAME','STUSPS')]
updated_column_names <- c('geoid','name','state')
setnames(states, updated_column_names)
states <- states[, category := 'state']
states <- states[, csa := 0]
setcolorder(states, final_column_order)

# Create a State FIPS Code to Abbreviation Lookup
state_lookup <- states[,c('geoid','state')]

# Counties
counties <- counties[,c('GEOID','NAME','STATEFP')]
counties <- merge(counties, state_lookup, by.x = "STATEFP", by.y = "geoid")
counties <- counties[,c('GEOID','NAME','state')]
updated_column_names <- c('geoid','name','state')
setnames(counties, updated_column_names)
counties <- counties[, category := 'county']
counties <- counties[, csa := 0]
setcolorder(counties, final_column_order)

# Statistical Areas
cbsa <- cbsa[,c('GEOID','NAME','CSAFP')]
cbsa[,c('CSAFP')] <- cbsa[, lapply(.SD, as.numeric), .SDcols = c('CSAFP')]
cbsa[is.na(cbsa)] <- 0
updated_column_names <- c('geoid','name','csa')
setnames(cbsa, updated_column_names)
cbsa <- cbsa[, category := 'statistical area']
cbsa$state <- gsub("^.*?, ","",cbsa$name)
setcolorder(cbsa, final_column_order)

# Urban Areas
urban_area <- urban_area[,c('GEOID10','NAME10')]
updated_column_names <- c('geoid','name')
setnames(urban_area, updated_column_names)
urban_area <- urban_area[, category := 'urban area']
urban_area <- urban_area[, csa := 0]
urban_area$state <- gsub("^.*?, ","",urban_area$name)
setcolorder(urban_area, final_column_order)

# Merge the various data.tables into a final clean table for inclusion in the central db
final_dt <- rbindlist(list(states, counties, cbsa, urban_area))
char_columns <- c('geoid','name','state','category')
final_dt[,char_columns] <- final_dt[, lapply(.SD, as.character), .SDcols = char_columns]
final_dt <- final_dt[, record_id := .I]
final_column_order <- c('record_id','geoid','name','state','category','csa')
setcolorder(final_dt, final_column_order)

# Now write data to the central database
dbWriteTable(elmer_connection, tablename, final_dt, overwrite=TRUE)

# Values for population download
source_python("census_population.py")
final_column_order <- c('record_id','geoid','year','value')
int_columns <- c('year','value')
data_years <- c('2010','2011','2012','2013','2014','2015','2016','2017')

# Statewide Population to Elmer
state_population <- create_population_table('state',data_years)
setDT(state_population)
state_population <- state_population[, record_id := .I]
setcolorder(state_population, final_column_order)
state_population[,int_columns] <- state_population[, lapply(.SD, as.integer), .SDcols = int_columns]
dbWriteTable(elmer_connection, 'national_statewide_population', state_population, overwrite=TRUE)

# County Population to Elmer
county_population <- create_population_table('county',data_years)
setDT(county_population)
county_population <- county_population[, record_id := .I]
setcolorder(county_population, final_column_order)
county_population[,int_columns] <- county_population[, lapply(.SD, as.integer), .SDcols = int_columns]
dbWriteTable(elmer_connection, 'national_county_population', county_population, overwrite=TRUE)

# MSA Population to Elmer
msa_population <- create_population_table('metropolitan statistical area/micropolitan statistical area',data_years)
setDT(msa_population)
msa_population <- msa_population[, record_id := .I]
setcolorder(msa_population, final_column_order)
msa_population[,int_columns] <- msa_population[, lapply(.SD, as.integer), .SDcols = int_columns]
dbWriteTable(elmer_connection, 'national_msa_population', msa_population, overwrite=TRUE)

# CSA Population to Elmer
csa_population <- create_population_table('combined statistical area',data_years)
setDT(csa_population)
csa_population <- csa_population[, record_id := .I]
setcolorder(csa_population, final_column_order)
csa_population[,int_columns] <- csa_population[, lapply(.SD, as.integer), .SDcols = int_columns]
dbWriteTable(elmer_connection, 'national_csa_population', csa_population, overwrite=TRUE)

# Urban Area Population to Elmer
ua_population <- create_population_table('urban area',data_years)
setDT(ua_population)
ua_population <- ua_population[, record_id := .I]
setcolorder(ua_population, final_column_order)
ua_population[,int_columns] <- ua_population[, lapply(.SD, as.integer), .SDcols = int_columns]
dbWriteTable(elmer_connection, 'national_urban_area_population', ua_population, overwrite=TRUE)

dbDisconnect(elmer_connection) 
```

```
```{r ofm_population_data, include=FALSE}
library(tidyverse)
library(readxl)
library(data.table)
library(lubridate)
library(odbc)
library(DBI)

ofm_url <- 'https://www.ofm.wa.gov/sites/default/files/public/legacy/pop/smallarea/data/xlsx/saep_bg10.xlsx'

# Details from OFM file including a list of tabs in the spreadsheet to iterate over
download.file(ofm_url, "working.xlsx", quiet = FALSE, mode = "wb")
ofm_pop_file <- paste0(getwd(),"/working.xlsx")

list_of_tabs <- excel_sheets(ofm_pop_file)
final_dt <- NULL

# Create a long form data.table of all OFM variables for inlcusion in Central DB
for (tabs in list_of_tabs) {
  ofm_pop_bg <- read_excel(ofm_pop_file,sheet = tabs, skip = 10)
  setDT(ofm_pop_bg)
  ofm_pop_bg <- na.omit(ofm_pop_bg)

  # Remove the columns that have deltas calculated
  cols_delta <- "([2][0-9]{3} to [2][0-9]{3})"
  cols_to_remove <- str_subset(colnames(ofm_pop_bg), cols_delta)
  ofm_pop_bg <- ofm_pop_bg[, (cols_to_remove):=NULL]

  # Drop Columns that are not the BlockGroup or Yearly Results
  cols_year <- "([2][0-9]{3})"
  cols_to_keep <- str_subset(colnames(ofm_pop_bg), cols_year)
  cols_to_keep <- c("Census Block Group Code Complete",cols_to_keep)
  ofm_pop_bg <- ofm_pop_bg[, ..cols_to_keep]

  # Now Rename columns to GEOID and Year before the data is transformed to long form
  cols_year <- "([2][0-9]{3})"
  updated_names <- str_subset(colnames(ofm_pop_bg), cols_year)
  updated_names <- str_sub(updated_names, -4, -1)
  updated_names <- c("geoid",updated_names)
  setnames(ofm_pop_bg,updated_names)

  working_dt <- melt(ofm_pop_bg, id.vars = c("geoid"), measure.vars = str_subset(colnames(ofm_pop_bg), cols_year))
  working_dt$date <- paste0("04/01/", working_dt$variable)
  working_dt <- working_dt[, date:=as.Date(date,"%d/%m/%Y")]
  working_dt$attribute <- tabs
  working_dt <- working_dt[, variable:=NULL]

  ifelse(is.null(final_dt), final_dt <- working_dt, final_dt <- rbindlist(list(final_dt, working_dt)))
}

# SQL Database Connection settings
elmer_connection <- dbConnect(odbc::odbc(),
  driver = "SQL Server",
  server = "sql2016\\DSADEV",
  database = "Sandbox",
  trusted_connection = "yes"
  )

dbWriteTable(elmer_connection, 'ofm_blockgroup_data', final_dt, overwrite=TRUE)
dbDisconnect(elmer_connection) 

# Now delete the temporary OFM downloaded file
file.remove(ofm_pop_file)

```