---
title: "Employment Trends"
output:
  html_document: default
---

Job growth is an indicator of a vibrant economy. In the Central Puget Sound Region (made up of King, Kitsap, Pierce and Snohomish counties), jobs are forecasted to increase from approximately 2 million today to over 3.2 million by the year 2050. An important part of regional planning is understanding how current growth trends align with future projections as well as how the job growth is distributed amongst the different parts of the region.

The Quarterly Census of Employment and Wages (QCEW) dataset is available for every metropolitan statistical area and county in the country and updates are released quarterly. This dataset includes both monthly estimates of total covered employment as well as average weekly wages by each geography. Covered Employment includes all jobs covered by unemployment insurance and captures roughly 90% of the total jobs in the region.

```{r include = FALSE}
library(odbc)
library(DBI)
library(data.table)
library(ggplot2)
library(scales)
library(knitr)
library(lubridate)
library(plotly)
library(leaflet)
library(sp)
library(rgdal)
library(DT)

# Genral FIPS ID's for standard summaries
psrc_region <- c('53033','53035', '53053', '53061')
king_county <- c('53033')
kitsap_county <- c('53035')
pierce_county <- c('53053')
snohomish_county <- c('53061')

# Function to Create a Bar Chart for a specific geograhpy
create_bar_chart <- function(table, county, xcolumn, ycolumn, fcolumn, ylimit) {

  psrc_colors <- c(
    "King County  " = "#91268F",
    "Kitsap County  " = "#F05A28",
    "Pierce County  " = "#8CC63E",
    "Snohomish County  " = "#00A7A0")

  # Create a Set of bar charts grouped by year for each area
  job_chart <- ggplot(table, aes(x = xcolumn, y= ycolumn, fill= fcolumn), show.legend = FALSE) +
  geom_col()+
  scale_fill_manual(values=psrc_colors)+
  scale_y_continuous(labels = comma, name = "Covered Employment",limits = c(0, ylimit))+
  xlab("Year")+
  theme_light()+
  ggtitle(paste0("Monthly Covered Employment: ", county))+
  theme(legend.position="none")

  job_chart <- ggplotly(job_chart)

}

# Function to create a County Specific Job set
create_geographic_summary <- function(table, geography, idate, pdate, cdate) {
  
  working_table <- table[area_fips %in% geography]
  job_totals <- working_table[,.(covered_employment=sum(value)),by=.(date)]

  initial_jobs <- job_totals[date %in% idate,covered_employment]
  previous_jobs <- job_totals[date %in% pdate,covered_employment] 
  current_jobs <- job_totals[date %in% cdate,covered_employment]

  total_change <- current_jobs - initial_jobs
  annual_change <- current_jobs - previous_jobs
  
  total_percent_change <- (total_change / initial_jobs) * 100
  annual_percent_change <- (annual_change / previous_jobs) * 100

  # Formatted numbers for text display
  i <- format(round(as.numeric(initial_jobs), -1), nsmall=0, big.mark=",")
  p <- format(round(as.numeric(previous_jobs), -1), nsmall=0, big.mark=",")
  c <- format(round(as.numeric(current_jobs), -1), nsmall=0, big.mark=",")

  tc <- format(round(as.numeric(total_change), -1), nsmall=0, big.mark=",")
  ac <- format(round(as.numeric(annual_change), -1), nsmall=0, big.mark=",")

  ptc <- format(round(as.numeric(total_percent_change), 1), nsmall=0, big.mark=",")
  pac <- format(round(as.numeric(annual_percent_change), 1), nsmall=0, big.mark=",")
  
  summary_results <- list(i, p, c, tc, ac, ptc, pac, total_change, annual_change)

}

# Table Name from the Central Database
working_dbtable <- "Craig.qcew_quarterly_msa_employment"

# SQL Database Connection settings
elmer_connection <- dbConnect(odbc::odbc(),
  driver = "SQL Server",
  server = "sql2016\\DSADEV",
  database = "Sandbox",
  trusted_connection = "yes"
  )

# Load the Table from the database and then close the connection
msa_data <- dbReadTable(elmer_connection,SQL(working_dbtable))
odbc::dbDisconnect(elmer_connection)

setDT(msa_data)

# convert the date column to info in format 'mm/dd/yyyy'
msa_data$date <- lubridate::mdy(msa_data$date)

# Figure out dates from the dataset for use in the summary
initial_date <- min(msa_data$date)
current_date <- max(msa_data$date)
previous_date <- as.Date(current_date) %m-% months(12)

initial_year <- year(initial_date)
previous_year <- year(previous_date)
current_year <- year(current_date)

first_month <- months(initial_date)
last_month <- months(current_date)

# Create Region specific datatables for further analysis
psrc_data <- msa_data[area_fips %in% psrc_region]
monthly_jobs <- psrc_data[attribute == 'jobs']
monthly_jobs$area_name <- gsub('Washington','',monthly_jobs$area_name)

```

## Regional Employment Changes

```{r include = FALSE}

# Create Summary Data by geography
region_results <- create_geographic_summary(monthly_jobs,psrc_region,initial_date,previous_date,current_date)

```

The economy of the Puget Sound region has been growing rapidly over the past five years, driven in large part by strong growth in our technology sectors. This job trend includes monthly employment data from `r toString(first_month)` of `r toString(initial_year)` to `r toString(last_month)` of `r toString(current_year)`. As noted above, these estimates are for total covered jobs within the four county region and include all public and private employment sectors.

In `r toString(first_month)` of `r toString(initial_year)` there were `r toString(region_results[[1]])` jobs in the Central Puget Sound region. By `r toString(last_month)` of `r toString(current_year)` the number of jobs in the region increased to `r toString(region_results[[3]])`, an increase of over `r toString(region_results[[4]])` jobs (a `r toString(region_results[[6]])`% change).

```{r echo = FALSE, warning= FALSE}
working_table <- monthly_jobs[area_fips %in% psrc_region]
region_jobs <- working_table[,.(covered_employment=sum(value)),by=.(date)]

regional_job_trend <- ggplot(region_jobs, aes(x=date,y=covered_employment)) +
  geom_line(color="#F05A28",size=1) +
  scale_y_continuous(labels = comma, name = "Covered Employment", limits = c(1000000, 3000000))+
  theme_light()+
  xlab("Year")+
  ggtitle("Monthly Covered Employment: PSRC Region")

regional_job_trend <- ggplotly(regional_job_trend)

regional_job_trend

```

In `r toString(last_month)` of `r toString(previous_year)` there were `r toString(region_results[[2]])` jobs in the Central Puget Sound region. By `r toString(last_month)` of `r toString(current_year)` the number of jobs in the region increased to `r toString(region_results[[3]])`, an increase of over `r toString(region_results[[5]])` jobs (a `r toString(region_results[[7]])`% change) in the past 12 months.


```{r echo = FALSE, warning= FALSE}

latest_month_jobs <- monthly_jobs[months(date) == last_month]
annual_jobs <- latest_month_jobs[,.(Region_Jobs=sum(value)),by=.(year)]
annual_jobs <- annual_jobs[order(year)]
annual_jobs$Change <- 0


for(i in 1:nrow(annual_jobs)) {
  if (i == 1) {
    annual_jobs$Change[i] <- '-'
    } else { annual_jobs$Change[i] <- (annual_jobs$Region_Jobs[i] - annual_jobs$Region_Jobs[i-1])}
}

annual_jobs$Region_Jobs <- format(round(as.numeric(annual_jobs$Region_Jobs), -1), nsmall=0, big.mark=",")
annual_jobs$Change <- format(round(as.numeric(annual_jobs$Change), -1), nsmall=0, big.mark=",")

datatable(annual_jobs, colnames = c('Year', 'Regional Employment', 'Annual Change'), caption = 'Table 1: Regional Covered Employment by Year') %>%
  formatStyle('year',`text-align` = 'center') %>%
  formatStyle('Region_Jobs',`text-align` = 'center') %>%
  formatStyle('Change',`text-align` = 'center')

```

## County Employment Changes {.tabset .tabset.fade .tabset-pills}

Jobs have increased all around the region since `r toString(initial_year)` but the level of growth has varied by county. King and Snohomish counties have accounted for a majority of the job growth since the end of the Great Recession but growth has accelerated in Kitsap and Pierce counties over the the past several years.

### King County
```{r include = FALSE}

# Create Summary Data by geography
king_results <- create_geographic_summary(monthly_jobs,king_county,initial_date,previous_date,current_date)
king_share_annual_change <- format(round((king_results[[9]] / region_results[[9]])*100,1), nsmall=0)
king_share_total_change <- format(round((king_results[[8]] / region_results[[8]])*100,1), nsmall=0)

```
In `r toString(last_month)` of `r toString(previous_year)` there were `r toString(king_results[[2]])` jobs in King County. By `r toString(last_month)` of `r toString(current_year)` the number of jobs in King County increased to `r toString(king_results[[3]])`, an increase of over `r toString(king_results[[5]])` jobs in the past 12 months. Over the past year, King County accounted for approximately `r toString(king_share_annual_change)`% of the total job growth in the four county region. Since `r toString(initial_year)`, King County has added over `r toString(king_results[[4]])` jobs which accounts for `r toString(king_share_total_change)`% of the total job growth in the four county region.

```{r echo = FALSE}
king_monthly_jobs <- monthly_jobs[area_fips %in% king_county]
king_jobs <- create_bar_chart(king_monthly_jobs, "King County", king_monthly_jobs$date, king_monthly_jobs$value, king_monthly_jobs$area_name, 2000000)
king_jobs
```

### Kitsap County
```{r include = FALSE}

# Create Summary Data by geography
kitsap_results <- create_geographic_summary(monthly_jobs,kitsap_county,initial_date,previous_date,current_date)
kitsap_share_annual_change <- format(round((kitsap_results[[9]] / region_results[[9]])*100,1), nsmall=0)
kitsap_share_total_change <- format(round((kitsap_results[[8]] / region_results[[8]])*100,1), nsmall=0)

```
In `r toString(last_month)` of `r toString(previous_year)` there were `r toString(kitsap_results[[2]])` jobs in Kitsap County. By `r toString(last_month)` of `r toString(current_year)` the number of jobs in Kitsap County increased to `r toString(kitsap_results[[3]])`, an increase of over `r toString(kitsap_results[[5]])` jobs in the past 12 months. Over the past year, Kitsap County accounted for approximately `r toString(kitsap_share_annual_change)`% of the total job growth in the four county region. Since `r toString(initial_year)`, Kitsap County has added over `r toString(kitsap_results[[4]])` jobs which accounts for `r toString(kitsap_share_total_change)`% of the total job growth in the four county region.

```{r echo = FALSE}
kitsap_monthly_jobs <- monthly_jobs[area_fips %in% kitsap_county]
kitsap_jobs <- create_bar_chart(kitsap_monthly_jobs, "Kitsap County", kitsap_monthly_jobs$date, kitsap_monthly_jobs$value, kitsap_monthly_jobs$area_name, 250000)
kitsap_jobs
```

### Pierce County
```{r include = FALSE}

# Create Summary Data by geography
pierce_results <- create_geographic_summary(monthly_jobs,pierce_county,initial_date,previous_date,current_date)
pierce_share_annual_change <- format(round((pierce_results[[9]] / region_results[[9]])*100,1), nsmall=0)
pierce_share_total_change <- format(round((pierce_results[[8]] / region_results[[8]])*100,1), nsmall=0)

```
In `r toString(last_month)` of `r toString(previous_year)` there were `r toString(pierce_results[[2]])` jobs in Pierce County. By `r toString(last_month)` of `r toString(current_year)` the number of jobs in Pierce County increased to `r toString(pierce_results[[3]])`, an increase of over `r toString(pierce_results[[5]])` jobs in the past 12 months. Over the past year, Pierce County accounted for approximately `r toString(pierce_share_annual_change)`% of the total job growth in the four county region. Since `r toString(initial_year)`, Pierce County has added over `r toString(pierce_results[[4]])` jobs which accounts for `r toString(pierce_share_total_change)`% of the total job growth in the four county region.

```{r echo = FALSE}
pierce_monthly_jobs <- monthly_jobs[area_fips %in% pierce_county]
pierce_jobs <- create_bar_chart(pierce_monthly_jobs, "Pierce County", pierce_monthly_jobs$date, pierce_monthly_jobs$value, pierce_monthly_jobs$area_name, 500000)
pierce_jobs
```

### Snohomish County
```{r include = FALSE}

# Create Summary Data by geography
snohomish_results <- create_geographic_summary(monthly_jobs,snohomish_county,initial_date,previous_date,current_date)
snohomish_share_annual_change <- format(round((snohomish_results[[9]] / region_results[[9]])*100,1), nsmall=0)
snohomish_share_total_change <- format(round((snohomish_results[[8]] / region_results[[8]])*100,1), nsmall=0)

```
In `r toString(last_month)` of `r toString(previous_year)` there were `r toString(snohomish_results[[2]])` jobs in Snohomish County. By `r toString(last_month)` of `r toString(current_year)` the number of jobs in Snohomish County increased to `r toString(snohomish_results[[3]])`, an increase of over `r toString(snohomish_results[[5]])` jobs in the past 12 months. Over the past year, Snohomish County accounted for approximately `r toString(snohomish_share_annual_change)`% of the total job growth in the four county region. Since `r toString(initial_year)`, Snohomish County has added over `r toString(snohomish_results[[4]])` jobs which accounts for `r toString(snohomish_share_total_change)`% of the total job growth in the four county region.

```{r echo = FALSE}
snohomish_monthly_jobs <- monthly_jobs[area_fips %in% snohomish_county]
snohomish_jobs <- create_bar_chart(snohomish_monthly_jobs, "Snohomish County", snohomish_monthly_jobs$date, snohomish_monthly_jobs$value, snohomish_monthly_jobs$area_name, 500000)
snohomish_jobs
```

## Statewide Job Growth {.tabset .tabset-fade .tabset-pills}

```{r include = FALSE}

# Create Region specific datatables for further analysis
washington_data <- msa_data[area_fips %in% grep("^530", area_fips, value = T) & area_name %in% grep("Washington$", area_name, value = T), ]
washington_data $area_fips <- gsub('^53','',washington_data $area_fips)

wa_monthly_jobs <- washington_data[attribute == 'jobs']

# Create a county level datatable of jobs from the previous year
wa_previous_year <- wa_monthly_jobs[date %in% previous_date]
wa_previous_year <- wa_previous_year[,.(area_fips,value)]
total_previous_jobs <- sum(wa_previous_year$value)
wa_previous_year$share <- (wa_previous_year$value / total_previous_jobs)*100
setnames(wa_previous_year,"value","previous_jobs")
setnames(wa_previous_year,"share","previous_share")
setkey(wa_previous_year,area_fips)

# Create a county level datatable of jobs from the current year
wa_current_year <- wa_monthly_jobs[date %in% current_date]
wa_current_year <- wa_current_year[,.(area_fips,value)]
total_current_jobs <- sum(wa_current_year$value)
wa_current_year$share <- (wa_current_year$value / total_current_jobs)*100
setnames(wa_current_year,"value","current_jobs")
setnames(wa_current_year,"share","current_share")
setkey(wa_current_year,area_fips)

# Join the previous and current job files before merging with the Shapefile
statewide_jobs <- merge(wa_previous_year,wa_current_year, all.x=TRUE)
statewide_jobs$delta <- statewide_jobs$current_jobs - statewide_jobs$previous_jobs
total_statewide_change <- sum(statewide_jobs$delta)
statewide_jobs$share <- (statewide_jobs$delta / total_statewide_change)*100

# Now merge the job data with the Washington State County Shapefile
zone.shape <- readOGR(dsn='D:/coding/employment/inputs',layer='countybackground_wgs84')
county_jobs <- sp::merge(zone.shape, statewide_jobs, by.x = "COUNTY_FIP", by.y = "area_fips")

```
### `r toString(previous_year)`
In `r toString(previous_year)`, there were a `r toString(total_previous_jobs)` jobs in Washington State.

```{r echo = FALSE}

#working_column <- "previous_share"
#test <- county_jobs[,get(eval(working_column))]

# Set bins for use in map based on range or attribute
rng <- range(county_jobs$previous_share)
max_bin <- max(abs(rng))
round_to <- 10^floor(log10(max_bin))
max_bin <- ceiling(max_bin/round_to)*round_to
breaks <- (sqrt(max_bin)*c(0.1, 0.2,0.4, 0.6, 0.8, 1))^2

bins <- c(0, breaks)
pal <- colorBin("Blues", domain = county_jobs$previous_share, bins = bins)

# Create a label format for the interactive map
labels <- paste0("<b>","County: ", "</b>",county_jobs$COUNTY_NM,
                 "<b> <br>","Jobs: ", "</b>", prettyNum(round(county_jobs$previous_jobs, -1), big.mark = ","),
                 "<b> <br>","Share: ", "</b>", prettyNum(round(county_jobs$previous_share, 1), big.mark = ","),"%") %>% lapply(htmltools::HTML)

county_map <- leaflet(data = county_jobs) %>% 
  addTiles() %>%
  addPolygons(fillColor = pal(county_jobs$previous_share),
              weight = 2,
              opacity = 1,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7,
              highlight = highlightOptions(
                weight =5,
                color = "76787A",
                dashArray ="",
                fillOpacity = 0.7,
                bringToFront = TRUE),
              label = labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"))%>%
  addLegend("bottomright", pal=pal, values = county_jobs$previous_share,
            title = "Share of Jobs",
            labFormat = labelFormat(suffix = "%"),
            opacity = 1)

county_map

```

### `r toString(current_year)`
In `r toString(current_year)`, there were even more jobs in the state .......

```{r echo = FALSE}

# Set bins for use in map based on range or attribute
rng <- range(county_jobs$current_share)
max_bin <- max(abs(rng))
round_to <- 10^floor(log10(max_bin))
max_bin <- ceiling(max_bin/round_to)*round_to
breaks <- (sqrt(max_bin)*c(0.1, 0.2,0.4, 0.6, 0.8, 1))^2

bins <- c(0, breaks)
pal <- colorBin("Blues", domain = county_jobs$current_share, bins = bins)

# Create a label format for the interactive map
labels <- paste0("<b>","County: ", "</b>",county_jobs$COUNTY_NM,
                 "<b> <br>","Jobs: ", "</b>", prettyNum(round(county_jobs$current_jobs, -1), big.mark = ","),
                 "<b> <br>","Share: ", "</b>", prettyNum(round(county_jobs$current_share, 1), big.mark = ","),"%") %>% lapply(htmltools::HTML)

county_map <- leaflet(data = county_jobs) %>% 
  addTiles() %>%
  addPolygons(fillColor = pal(county_jobs$current_share),
              weight = 2,
              opacity = 1,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7,
              highlight = highlightOptions(
                weight =5,
                color = "76787A",
                dashArray ="",
                fillOpacity = 0.7,
                bringToFront = TRUE),
              label = labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"))%>%
  addLegend("bottomright", pal=pal, values = county_jobs$current_share,
            title = "Share of Jobs",
            labFormat = labelFormat(suffix = "%"),
            opacity = 1)

county_map

```
### `r toString(previous_year)` to `r toString(current_year)`
In `r toString(current_year)`, there were even more jobs in the state .......

```{r echo = FALSE}

# Set bins for use in map based on range or attribute
rng <- range(county_jobs$current_share)
max_bin <- max(abs(rng))
round_to <- 10^floor(log10(max_bin))
max_bin <- ceiling(max_bin/round_to)*round_to
breaks <- (sqrt(max_bin)*c(0.1, 0.2,0.4, 0.6, 0.8, 1))^2

bins <- c(0, breaks)
pal <- colorBin("Blues", domain = county_jobs$current_share, bins = bins)

# Create a label format for the interactive map
labels <- paste0("<b>","County: ", "</b>",county_jobs$COUNTY_NM,
                 "<b> <br>","Jobs: ", "</b>", prettyNum(round(county_jobs$current_jobs, -1), big.mark = ","),
                 "<b> <br>","Share: ", "</b>", prettyNum(round(county_jobs$current_share, 1), big.mark = ","),"%") %>% lapply(htmltools::HTML)

county_map <- leaflet(data = county_jobs) %>% 
  addTiles() %>%
  addPolygons(fillColor = pal(county_jobs$current_share),
              weight = 2,
              opacity = 1,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7,
              highlight = highlightOptions(
                weight =5,
                color = "76787A",
                dashArray ="",
                fillOpacity = 0.7,
                bringToFront = TRUE),
              label = labels,
              labelOptions = labelOptions(
                style = list("font-weight" = "normal", padding = "3px 8px"),
                textsize = "15px",
                direction = "auto"))%>%
  addLegend("bottomright", pal=pal, values = county_jobs$current_share,
            title = "Share of Jobs",
            labFormat = labelFormat(suffix = "%"),
            opacity = 1)

county_map

```
# National Comparison
