---
title: "Mapping Attributes: Housing Type and Sales"
output: html_notebook
---


```{r}
wrk.dir <- "C:/Users/CLam/Desktop/data-science/assessor"
curr.dir <- getwd()
setwd(wrk.dir)

source("price_functions.R")

# assemble all tables into list
dlist <- read.assessor.files("Kitsap", "2018", "dwnld_2018_11_06")
# tidy data
colnames(dlist$Residential_5_yr_dwelling_sales)[grep('\"', colnames(dlist$Residential_5_yr_dwelling_sales))] <- c("REET_no", "Living_area")
dlist$Residential_5_yr_dwelling_sales$REET_no <- str_replace(dlist$Residential_5_yr_dwelling_sales$REET_no, '\"', "")
dlist$Residential_5_yr_dwelling_sales$Living_area <- str_replace(dlist$Residential_5_yr_dwelling_sales$Living_area, '\"', "")
```

#### General Notes 

* https://www.kitsapgov.com/assessor/Pages/DataDownload.aspx
* Sales data is publicly available only for the last 5 years  
* rp_acct_id (aka LRSN) is the primary key

#### Tables::Columns 

* Residential_5_yr_dwelling_sales  
    + REET_no, Sale_Dt, Price, Acct_no, LRSN, Dwellings, House_type
* Parcels  
    + rp_acct_id, acct_no, property_class
* Codes (for Parcels::property_class)

```{r}
# preview all columns in all tables
print.table.structure(dlist)
```

#### Housing Type Descriptions  

* Housing type descriptions from Residential_5_yr_dwelling_sales are similiar to those found in Dwellings with a few additional values (likely MH types from MH table)
```{r}
# Compare housing type descriptions
unique(dlist$Residential_5_yr_dwelling_sales$House_type)
unique(dlist$Dwellings$house_type)

```
```{r}
assemble.housing.price.dataset <- function() {
  sales <- dlist$Residential_5_yr_dwelling_sales[, .(REET_no, Sale_Dt, Price, Acct_no, LRSN, Dwellings, House_type)]
  prcl <- dlist$Parcels[, .(rp_acct_id, acct_no, property_class)]
  prcl[dlist$codes[Field_name == "property_class"], on = c("property_class" = "Code"), property_class_description := i.Description]
  sales[prcl, on = c("LRSN" = "rp_acct_id", "Acct_no" = "acct_no"), property_class_description := i.property_class_description]
}

df <- assemble.housing.price.dataset()

```

