---
title: "Calculating trip rates"
author: "Polina Butrina"
date: "10/11/2020"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(data.table)
library(tidyverse)
library(DT)
library(openxlsx)
library(odbc)
library(DBI)
library(psych)

p_MOE <- 0.5
z<-1.645
missing_codes <- c('Missing: Technical Error', 'Missing: Non-response', 
                   'Missing: Skip logic', 'Children or missing', 'Prefer not to answer')

# connecting to Elmer
db.connect <- function() {
  elmer_connection <- dbConnect(odbc(),
                                driver = "SQL Server",
                                server = "AWS-PROD-SQL\\Sockeye",
                                database = "Elmer",
                                trusted_connection = "yes"
  )
}

# a function to read tables and queries from Elmer
read.dt <- function(astring, type =c('table_name', 'sqlquery')) {
  elmer_connection <- db.connect()
  if (type == 'table_name') {
    dtelm <- dbReadTable(elmer_connection, SQL(astring))
  } else {
    dtelm <- dbGetQuery(elmer_connection, SQL(astring))
  }
  dbDisconnect(elmer_connection)
  dtelm
}

create_table_one_var = function(var1, table_temp,table_type ) {
  #table_temp = recategorize_var_upd(var2,table_temp)
  #print(table_temp)
  if (table_type == "household" | table_type == "person" ) {
    weight_2017 = "hh_wt_revised"
    weight_2019 = "hh_wt_2019"
    weight_comb = "hh_wt_combined"
  } else if (table_type == "trip") {
    weight_2017 = "trip_weight_revised"
    weight_2019 = "trip_wt_2019"
    weight_comb = "trip_wt_combined"  
  } 
  
  temp = table_temp %>% select(!!sym(var1), all_of(weight_2017), all_of(weight_2019), all_of(weight_comb)) %>% 
    filter(!.[[1]] %in% missing_codes, !is.na(.[[1]])) %>% 
    group_by(!!sym(var1)) %>% 
    summarise(n=n(),sum_wt_comb = sum(.data[[weight_comb]],na.rm = TRUE),sum_wt_2017 = sum(.data[[weight_2017]],na.rm = TRUE),sum_wt_2019 = sum(.data[[weight_2019]],na.rm = TRUE)) %>% 
    mutate(perc_comb = sum_wt_comb/sum(sum_wt_comb)*100, perc_2017 = sum_wt_2017/sum(sum_wt_2017)*100, perc_2019 = sum_wt_2019/sum(sum_wt_2019)*100,delta = perc_2019-perc_2017) %>% 
    ungroup() %>%  mutate(MOE=1.65*(0.25/sum(n))^(1/2)*100) %>% arrange(desc(perc_comb))
  return(temp)
}

```
```{r, include=FALSE}
options(tinytex.verbose = TRUE)
```


## R Markdown

For general HHTS data analysis tips and tricks, please refer to this intro from RSG.

There are several analyses you can do with trips:
•	Find trip rates
•	Find a number of trips

### Find Trip Rates
Trips rates (or the number of trips per day among groups) can be found using following steps:

1.	Filter trip table to just shopping trips


```{r cars}
trip.query = paste("SELECT * FROM HHSurvey.v_trips_2017_2019")
trips = read.dt(trip.query, 'sqlquery')

shop_trips <- trips %>% filter(d_purp_cat=='Shop')

```

2.	Sum trip_wt_combined multiplied 

```{r}
sum(shop_trips$trip_wt_combined)

shop_trips_gender<-create_table_one_var('gender', shop_trips, 'trip')

shop_trips_gender

```

3.	Merge trip counts with the day table

```{r}
days<-paste("SELECT * FROM HHSurvey.v_days_2017_2019_in_house")
person_days=read.dt(days, 'sqlquery')

person.query = paste("SELECT * FROM HHSurvey.v_persons_2017_2019")
person = read.dt(person.query, 'sqlquery')

person_days_2<-merge(person_days, person, by.x='personid', by.y='person_dim_id')

```

4.	Summarize day trips by gender

```{r}

person_day_gender <- person_days_2 %>% group_by(gender) %>%
                      summarise(n=n(), day_combined = sum(hh_day_wt_combined.x))

person_day_gender

```

5.	Calculate trip rate as sum of shopping trips divided by the number of weighted person days

```{r}

shop_trips_3 <- merge(shop_trips_gender, person_day_gender, by = 'gender')
shop_trips_3 %>% mutate(trip_rate = sum_wt_comb/day_combined) %>% select(gender,trip_rate)

```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
