---
title: "Puget Sound Trends"
author: "Executive Board Presentation &#13; Puget Sound Regional Council"
date: "January 24, 2019"
output: 
  powerpoint_presentation:
    reference_doc: psrc-template.pptx
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(odbc)
library(DBI)
library(data.table)
library(ggplot2)
library(scales)
library(knitr)
library(lubridate)
library(plotly)
library(leaflet)
library(sp)
library(rgdal)
library(DT)

# Genral FIPS ID's for standard summaries
psrc_region <- c('53033','53035', '53053', '53061')
king_county <- c('53033')
kitsap_county <- c('53035')
pierce_county <- c('53053')
snohomish_county <- c('53061')

psrc_colors <- c("King County  " = "#91268F",
                 "Kitsap County  " = "#F05A28",
                 "Pierce County  " = "#8CC63E",
                 "Snohomish County  " = "#00A7A0")

# Function to Create a Bar Chart for a specific geograhpy
create_bar_chart <- function(table, county, xcolumn, ycolumn, fcolumn, ylimit) {

  psrc_colors <- c(
    "King County  " = "#91268F",
    "Kitsap County  " = "#F05A28",
    "Pierce County  " = "#8CC63E",
    "Snohomish County  " = "#00A7A0")

  # Create a Set of bar charts grouped by year for each area
  job_chart <- ggplot(table, aes(x = xcolumn, y= ycolumn, fill= fcolumn), show.legend = FALSE) +
  geom_col()+
  scale_fill_manual(values=psrc_colors)+
  scale_y_continuous(labels = comma, name = "Covered Employment",limits = c(0, ylimit))+
  xlab("Year")+
  theme_light()+
  ggtitle(paste0("Monthly Covered Employment: ", county))+
  theme(legend.position="none")

}

# Function to create a County Specific Job set
create_geographic_summary <- function(table, geography, idate, pdate, cdate) {
  
  working_table <- table[area_fips %in% geography]
  job_totals <- working_table[,.(covered_employment=sum(value)),by=.(date)]

  initial_jobs <- job_totals[date %in% idate,covered_employment]
  previous_jobs <- job_totals[date %in% pdate,covered_employment] 
  current_jobs <- job_totals[date %in% cdate,covered_employment]

  total_change <- current_jobs - initial_jobs
  annual_change <- current_jobs - previous_jobs
  
  total_percent_change <- (total_change / initial_jobs) * 100
  annual_percent_change <- (annual_change / previous_jobs) * 100

  # Formatted numbers for text display
  i <- format(round(as.numeric(initial_jobs), -1), nsmall=0, big.mark=",")
  p <- format(round(as.numeric(previous_jobs), -1), nsmall=0, big.mark=",")
  c <- format(round(as.numeric(current_jobs), -1), nsmall=0, big.mark=",")

  tc <- format(round(as.numeric(total_change), -1), nsmall=0, big.mark=",")
  ac <- format(round(as.numeric(annual_change), -1), nsmall=0, big.mark=",")

  ptc <- format(round(as.numeric(total_percent_change), 1), nsmall=0, big.mark=",")
  pac <- format(round(as.numeric(annual_percent_change), 1), nsmall=0, big.mark=",")
  
  summary_results <- list(i, p, c, tc, ac, ptc, pac, total_change, annual_change)

}

# SQL Database Connection settings
elmer_connection <- dbConnect(odbc::odbc(),
  driver = "SQL Server",
  server = "sql2016\\DSADEV",
  database = "Sandbox",
  trusted_connection = "yes"
  )
```

## Overview

Topics covered in today's presentation:

- How many jobs are there?
- Where do people work?
- SeaTac Airport passenger volume changes

# Employment & Wage Trends

```{r job_growth_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Table Name from the Central Database
working_dbtable <- "Craig.qcew_quarterly_msa_employment"

# Load the Table from the database and then close the connection
msa_data <- dbReadTable(elmer_connection,SQL(working_dbtable))
odbc::dbDisconnect(elmer_connection)

setDT(msa_data)

# convert the date column to info in format 'mm/dd/yyyy'
msa_data$date <- lubridate::mdy(msa_data$date)

# Figure out dates from the dataset for use in the summary
initial_date <- min(msa_data$date)
current_date <- max(msa_data$date)
previous_date <- as.Date(current_date) %m-% months(12)

initial_year <- year(initial_date)
previous_year <- year(previous_date)
current_year <- year(current_date)

first_month <- months(initial_date)
last_month <- months(current_date)

# Create Region specific datatables for further analysis
psrc_data <- msa_data[area_fips %in% psrc_region]
monthly_jobs <- psrc_data[attribute == 'jobs']
monthly_jobs$area_name <- gsub('Washington','',monthly_jobs$area_name)

working_table <- monthly_jobs[area_fips %in% psrc_region]
region_jobs <- working_table[,.(covered_employment=sum(value)),by=.(date)]

# Create a line graph for the regional job trend
regional_job_trend <- ggplot(region_jobs, aes(x=date,y=covered_employment)) +
  geom_line(color="#F05A28",size=1.5) +
  scale_y_continuous(labels = comma, name = "Covered Employment", limits = c(1000000, 3000000))+
  theme_light()+
  theme(panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"))+
  xlab("Year")+
  ggtitle("Monthly Covered Employment: PSRC Region")

# Create Summary Data by geography
region_results <- create_geographic_summary(monthly_jobs,psrc_region,initial_date,previous_date,current_date)

king_results <- create_geographic_summary(monthly_jobs,king_county,initial_date,previous_date,current_date)
king_share_annual_change <- format(round((king_results[[9]] / region_results[[9]])*100,1), nsmall=0)
king_share_total_change <- format(round((king_results[[8]] / region_results[[8]])*100,1), nsmall=0)

kitsap_results <- create_geographic_summary(monthly_jobs,kitsap_county,initial_date,previous_date,current_date)
kitsap_share_annual_change <- format(round((kitsap_results[[9]] / region_results[[9]])*100,1), nsmall=0)
kitsap_share_total_change <- format(round((kitsap_results[[8]] / region_results[[8]])*100,1), nsmall=0)

pierce_results <- create_geographic_summary(monthly_jobs,pierce_county,initial_date,previous_date,current_date)
pierce_share_annual_change <- format(round((pierce_results[[9]] / region_results[[9]])*100,1), nsmall=0)
pierce_share_total_change <- format(round((pierce_results[[8]] / region_results[[8]])*100,1), nsmall=0)

snohomish_results <- create_geographic_summary(monthly_jobs,snohomish_county,initial_date,previous_date,current_date)
snohomish_share_annual_change <- format(round((snohomish_results[[9]] / region_results[[9]])*100,1), nsmall=0)
snohomish_share_total_change <- format(round((snohomish_results[[8]] / region_results[[8]])*100,1), nsmall=0)

```

## Regional Employment Continues to Increase

:::::::::::::: {.columns}
::: {.column}
In `r toString(last_month)` of `r toString(current_year)` there were `r toString(region_results[[3]])` in the region, an increase of over `r toString(region_results[[4]])` jobs (a `r toString(region_results[[6]])`% change) since `r toString(initial_year)`.

- King County has added over `r toString(king_results[[4]])` jobs (`r toString(king_share_total_change)`% of the regional increase).

- Kitsap County has added over `r toString(kitsap_results[[4]])` jobs (`r toString(kitsap_share_total_change)`% of the regional increase).

- Pierce County has added over `r toString(pierce_results[[4]])` jobs (`r toString(pierce_share_total_change)`% of the regional increase).

- Snohomish County has added over `r toString(snohomish_results[[4]])` jobs (`r toString(snohomish_share_total_change)`% of the regional increase).

:::
::: {.column}
```{r regional_job_trend, echo = FALSE}
regional_job_trend
```
:::
::::::::::::::

```{r wage_growth_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Create Region specific datatables for further analysis
weekly_wages <- psrc_data[attribute == 'wages']
weekly_wages$area_name <- gsub('Washington','',weekly_wages$area_name)
weekly_wages <- weekly_wages[months(date) == last_month]

# King County
king_wage_initial <- weekly_wages[year %in% initial_year & area_fips=='53033',value]
king_wage_current <- weekly_wages[year %in% current_year & area_fips=='53033',value]
king_wage_delta <- king_wage_current - king_wage_initial
king_wage_percentage <- format(round((king_wage_delta / king_wage_initial)*100,1), nsmall=0)
ikinw <- format(round(as.numeric(king_wage_initial), -1), nsmall=0, big.mark=",")
ckinw <- format(round(as.numeric(king_wage_current), -1), nsmall=0, big.mark=",")

# Kitsap County
kitsap_wage_initial <- weekly_wages[year %in% initial_year & area_fips=='53035',value]
kitsap_wage_current <- weekly_wages[year %in% current_year & area_fips=='53035',value]
kitsap_wage_delta <- kitsap_wage_current - kitsap_wage_initial
kitsap_wage_percentage <- format(round((kitsap_wage_delta / kitsap_wage_initial)*100,1), nsmall=0)
ikitw <- format(round(as.numeric(kitsap_wage_initial), -1), nsmall=0, big.mark=",")
ckitw <- format(round(as.numeric(kitsap_wage_current), -1), nsmall=0, big.mark=",")

# pierce County
pierce_wage_initial <- weekly_wages[year %in% initial_year & area_fips=='53053',value]
pierce_wage_current <- weekly_wages[year %in% current_year & area_fips=='53053',value]
pierce_wage_delta <- pierce_wage_current - pierce_wage_initial
pierce_wage_percentage <- format(round((pierce_wage_delta / pierce_wage_initial)*100,1), nsmall=0)
ipiew <- format(round(as.numeric(pierce_wage_initial), -1), nsmall=0, big.mark=",")
cpiew <- format(round(as.numeric(pierce_wage_current), -1), nsmall=0, big.mark=",")

# snohomish County
snohomish_wage_initial <- weekly_wages[year %in% initial_year & area_fips=='53061',value]
snohomish_wage_current <- weekly_wages[year %in% current_year & area_fips=='53061',value]
snohomish_wage_delta <- snohomish_wage_current - snohomish_wage_initial
snohomish_wage_percentage <- format(round((snohomish_wage_delta / snohomish_wage_initial)*100,1), nsmall=0)
isnow <- format(round(as.numeric(snohomish_wage_initial), -1), nsmall=0, big.mark=",")
csnow <- format(round(as.numeric(snohomish_wage_current), -1), nsmall=0, big.mark=",")

# Create a line chart for wagres by county
county_wage_trend <- ggplot(weekly_wages, aes(x=date,y=value, color=area_name)) +
  geom_line(size = 1.5) +
  scale_color_manual(values=psrc_colors)+
  scale_y_continuous(labels = comma, name = "Weekly Wage", limits = c(0, 2500))+
  theme_light()+
  theme(panel.border = element_blank(),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.line = element_line(colour = "black"))+
  theme(legend.position="bottom")+
  theme(legend.title=element_blank())+
  xlab("Year")+
  ggtitle("Weekly Wages by County: PSRC Region")


```

## Wages are increasing - but at different rates

:::::::::::::: {.columns}
::: {.column}
Between `r toString(last_month)` of `r toString(initial_year)` and `r toString(last_month)` of `r toString(current_year)`, wages have been growing at different rates across the four counties.

- Average weekly wages in King County changed from $`r toString(ikinw)` to $`r toString(ckinw)` (a `r toString(king_wage_percentage)`% change).

- Average weekly wages in Kitsap County changed from $`r toString(ikitw)` to $`r toString(ckitw)` (a `r toString(kitsap_wage_percentage)`% change).

- Average weekly wages in Pierce County changed from $`r toString(ipiew)` to $`r toString(cpiew)` (a `r toString(pierce_wage_percentage)`% change).

- Average weekly wages in Snohomish County changed from $`r toString(isnow)` to $`r toString(csnow)` (a `r toString(snohomish_wage_percentage)`% change).

:::
::: {.column}
```{r county_wage_trend, echo = FALSE}
county_wage_trend
```

:::
::::::::::::::

# Where do people work?

```{r worker_flow_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# SQL Database Connection settings
elmer_connection <- dbConnect(odbc::odbc(),
  driver = "SQL Server",
  server = "sql2016\\DSADEV",
  database = "Sandbox",
  trusted_connection = "yes"
  )

# Table Name from the Central Database
working_dbtable <- "lehd_otm_county_workers"

# Load the Table from the database and then close the connection
lehd_data <- dbReadTable(elmer_connection,SQL(working_dbtable))
odbc::dbDisconnect(elmer_connection)

setDT(lehd_data)

```

## Where King County Residents Work

```{r king_worker_flow_setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

working_geography <- c('033')
working_years <- c('2005')
working_table <- lehd_data[home_county %in% working_geography]
working_table <- working_table[year %in% working_years]

# Create a summary of Jobs by Jurisdiction
total_workers <- working_table[,.(number_of_workers=sum(value)),by=.(work_county,work_city)]
total_workers <- total_workers[order(-rank(number_of_workers))]
nms <- c("County", "Jurisdiction", "Workers")
setnames(total_workers, nms)
total_workers$Workers <- format(round(as.numeric(total_workers$Workers), -1), nsmall=0, big.mark=",")

top_ten <- total_workers[1:10]
top_ten_commute_cities <- kable(top_ten)

# Merge the job data with the Washington State County Shapefile
jobs_to_merge <- working_table[,c("work_block","value")]
block.shape <- readOGR(dsn='D:/coding/data-science/trends/input/blocks',layer='wa_blocks_wgs1984',stringsAsFactors = FALSE)
king_jobs <- sp::merge(block.shape, jobs_to_merge, by.x = "GEOID10", by.y = "work_block")
king_jobs$value[is.na(king_jobs$value)] <- 0

# Set bins for use in map based on range
rng <- range(king_jobs$value)
max_bin <- max(abs(rng))
round_to <- 10^floor(log10(max_bin))
max_bin <- ceiling(max_bin/round_to)*round_to
breaks <- (sqrt(max_bin)*c(0.1, 0.2,0.4, 0.6, 0.8, 1))^2

bins <- c(0, breaks)
pal <- colorBin("Blues", domain = king_jobs$value, bins = bins)

king_worker_cities <- leaflet(data = king_jobs) %>% 
  addTiles() %>%
  addPolygons(fillColor = pal(king_jobs$value_share),
              weight = 2,
              opacity = 1,
              color = "white",
              dashArray = "3",
              fillOpacity = 0.7)%>%
  addLegend("bottomright", pal=pal, values = king_jobs$value,
            title = "Number of Workers",
            labFormat = labelFormat(suffix = " workers"),
            opacity = 1)

```

:::::::::::::: {.columns}
::: {.column}

```{r king_workers_trend, echo = FALSE}
top_ten_commute_cities
```

:::
::: {.column}

```{r king_workers_cities, echo = FALSE}
#king_worker_cities
```

:::
::::::::::::::

## R Markdown

This is an R Markdown presentation. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document.
