---
title: "Comparing the percent of single family zoning in a block group to the dissimilarity index"
author: "suzanne"
date: "9/21/2021"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Dissimilarity indices vs share of renters


```{r echo=FALSE, message=FALSE, warning=FALSE}
setwd("~/GitHub/data-science/dissimilarity")

library(broom)
library(dplyr)
library(corrr)
library(ggplot2)
library(tidyverse)
library(Hmisc)
library(DBI)
library(psrccensus)
library(odbc)
library(foreign)
res_diss<-read.dbf('diss_res.dbf')
```

## Spearman correlation between dissimilarity indices and % renters.

```{r echo=FALSE, message=FALSE, warning=FALSE}
census_df<-read.csv('census_tract_vars.csv')
census_df$GEOID<-as.character(census_df$GEOID)
dissim_names<-c('White__Bla', 'White_AIAN', 'White__API',	 'White__His',	'White__Min',	'Black__AIA',	'Black__API',	'Black__Oth','Black__His',	'AIAN__Asia',	'AIAN__Othe','AIAN__Hisp',	'API__Other' ,	'API__Hispa',	'Other_2___'
)
yvar <- names(res_diss)[names(res_diss) %in% dissim_names]
res_diss$GEOID<-str_sub(as.character(res_diss$geoid10),1,11)
res_diss_census<-merge(res_diss, census_df, on='GEOID')
xvar<-'rent'

yrbuilt<-get_acs_recs(geography = 'tract',
             table.names = c('B25038'),
             years=c(2019),
             acs.type = 'acs5')
yrbuilt$GEOID<-as.character(yrbuilt$GEOID)

res_diss_census2<-merge(res_diss_census, median_yrbuilt, on='GEOID')
res_diss_census2<-res_diss_census2 %>% filter(estimate>0)

# sig_test<-lapply(yvar,
#        function(yvar, xvar, res_diss)
#        {
#          ggplot(res_diss_census, aes(x=xvar,y=yvar))+geom_point()
#          cor.test(res_diss[[xvar]], res_diss[[yvar]], method='spearman') %>%
#            tidy()
#        },
#        xvar,
#        res_diss) %>%
#   bind_rows() %>%mutate('dissimiliarity_index'=dissim_names)
# sig_test
# write.table(sig_test, "clipboard", sep='\t', row.names=FALSE)

ggplot(res_diss_census, aes(x=rent,y=White__Bla))+geom_point()+geom_smooth(method='lm')
       
cor.test(res_diss_census$rent, res_diss_census$White__Bla, method='spearman') %>%
           tidy()

summary(lm(White__Bla~rent, data=res_diss_census))

cor.test(census_df$rent, census_df$nonwhite, method='spearman')

ggplot(census_df, aes(x=nonwhite,y=rent))+geom_point()+geom_smooth(method='lm')


ggplot(res_diss_census2, aes(x=nonwhite,y=estimate))+geom_point()+geom_smooth(method='lm')

ggplot(res_diss_census2, aes(x=cost_burdened,y=estimate))+geom_point()+geom_smooth(method='lm')
summary(lm(White__Bla~rent+estimate, data=res_diss_census2))

```

## Pearson correlation between dissimilarity indexes and % single family zoning.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# sig_test_pearson<-lapply(yvar,
#        function(yvar, xvar, res_diss)
#        {
#          cor.test(res_diss[[xvar]], res_diss[[yvar]], method='pearson') %>%
#            tidy()
#        },
#        xvar,
#        res_diss) %>%
# ```


```{r echo=FALSE, message=FALSE, warning=FALSE}
bind_rows()%>%mutate('dissimiliarity_index'=dissim_names)%>%select(-c(statistic, conf.low, conf.high))

sig_test_pearson
write.table(sig_test_pearson, "clipboard", sep='\t', row.names=FALSE)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
db.connect <- function() {
  elmer_connection <- dbConnect(odbc(),
                                driver = "SQL Server",
                                server = "AWS-PROD-SQL\\Sockeye",
                                database = "Elmer",
                                trusted_connection = "yes"
  )
}

# a function to read tables and queries from Elmer
read.dt <- function(astring, type =c('table_name', 'sqlquery')) {
  elmer_connection <- db.connect()
  if (type == 'table_name') {
    dtelm <- dbReadTable(elmer_connection, SQL(astring))
  } else {
    dtelm <- dbGetQuery(elmer_connection, SQL(astring))
  }
  dbDisconnect(elmer_connection)
  dtelm
}

block_group_city<-read.dt("select pd.block_group_geoid10 from small_areas.parcel_dim pd where pd.city_name_2020 = 'Seattle' group by pd.block_group_geoid10 order by pd.block_group_geoid10", 'sqlquery')

block_group_city <- block_group_city %>% mutate(block_group_geoid10 =as.character(block_group_geoid10))
res_diss<-res_diss %>% mutate(geoid10=as.character(geoid10))
res_dissim_seattle<- dplyr::left_join(block_group_city ,res_diss, by = c("block_group_geoid10"="geoid10"))

```
## Spearman correlation between % single family zoning and dissimilarity for Seattle only
```{r echo=FALSE, message=FALSE, warning=FALSE}
sig_test<-lapply(yvar,
       function(yvar, xvar, res_diss)
       {
   
         cor.test(res_dissim_seattle[[xvar]], res_dissim_seattle[[yvar]], method='spearman') %>%
           tidy()
       },
       xvar,
       res_dissim_seattle) %>%
  bind_rows() %>%mutate('dissimiliarity_index'=dissim_names)

sig_test
write.table(sig_test, "clipboard", sep='\t', row.names=FALSE)


```
## Pearson correlation between % single family zoning and dissimilarity for Seattle only

```{r echo=FALSE, message=FALSE, warning=FALSE}
sig_test_pearson<-lapply(yvar,
       function(yvar, xvar, res_diss)
       {
         cor.test(res_dissim_seattle[[xvar]], res_dissim_seattle[[yvar]], method='pearson') %>%
           tidy()
       },
       xvar,
       res_diss_seattle) %>%
  bind_rows()%>%mutate('dissimiliarity_index'=dissim_names)%>%select(-c(statistic, conf.low, conf.high))
sig_test_pearson
write.table(sig_test_pearson, "clipboard", sep='\t', row.names=FALSE)
```

This is a summary of the the range of the percent single family variable. There are many block groups with 100% single family zoning, and many with 0% single family zoning. Higher shares of single family zoning are more common, such as greater than 75%, than lower shares such as 25%

Histogram of the % single family zoning

```{r echo=FALSE, message=FALSE, warning=FALSE}
ggplot(res_diss, aes(x=prct_sf)) +geom_histogram()
```

In this summary, I am plotting histograms for the values taken on by each dissimilarity index. The distributions of each dissimilarity index can tell a story. I think we need to work together to figure out what it means.

## Histograms for each dissimilarity index

```{r echo=FALSE, message=FALSE, warning=FALSE}
for(name in dissim_names){
print(ggplot(res_diss, aes_string(x=name)) + geom_histogram())
}

```

## Scatterplots between % single family zoning and the dissimilarity index

```{r echo=FALSE, message=FALSE, warning=FALSE}
for(name in dissim_names){
print(ggplot(res_diss, aes_string(x='prct_sf', y=name))+
  geom_point()+
  geom_smooth(method=lm, formula='y ~poly(x,3)'))
}
```
```{r}
res_diss_seattle<-res_diss_seattle %>% mutate(
  sf_or_not= case_when(
    prct_sf<1~ 'Less than 100% single family zoning',
    prct_sf==1 ~ '100% single family zoning'
  ))
res_diss_seattle<-res_diss_seattle %>% mutate(
  sf_or_not= case_when(
    prct_sf<1~ 'Less than 100% single family zoning',
    prct_sf==1 ~ '100% single family zoning'
  ))

res_diss_grouped<-res_diss_seattle %>%group_by(sf_or_not)%>%summarize_at(dissim_names, mean, na.rm=TRUE)
write.table(res_diss_grouped, 'clipboard', sep='\t', row.names=FALSE )
```


```{r}
res_diss<-res_diss %>% mutate(
  sf_or_not= case_when(
    prct_sf<1~ 'Less than 100% single family zoning',
    prct_sf==1 ~ '100% single family zoning'
  ))
```
```{r}
res_diss_grouped<-res_diss %>%group_by(sf_or_not)%>%summarize_at(dissim_names, mean, na.rm=TRUE)
write.table(res_diss_grouped, 'clipboard', sep='\t', row.names=FALSE )
```


```{r}

```{r}
res_diss_grouped<-res_diss %>%group_by(sf_or_not)%>%summarize_at(dissim_names, mean, na.rm=TRUE)
write.table(res_diss_grouped, 'clipboard', sep='\t', row.names=FALSE )

```


```
```{r}

res_diss_seattle<-res_diss_seattle %>% mutate(
  sf_or_not= case_when(
    prct_sf<1~ 'Less than 100% single family zoning',
    prct_sf==1 ~ '100% single family zoning'
  ))
res_diss_seattle<-res_diss_seattle %>% mutate(
  sf_or_not= case_when(
    prct_sf<1~ 'Less than 100% single family zoning',
    prct_sf==1 ~ '100% single family zoning'
  ))
```{r}

```

res_diss_grouped<-res_diss_seattle %>%group_by(sf_or_not)%>%summarize_at(dissim_names, mean, na.rm=TRUE)
write.table(res_diss_grouped, 'clipboard', sep='\t', row.names=FALSE )
```
