---
title: "Analyzing the Dissimilarity Indices for the PSRC region"
author: "Data Science"
date: "June 2021"
output:
  html_document:
    df_print: paged
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Does Single Family Zoning correlate with greater segregation?

The dissimilarity index was calculated at a block group level. The Future Land Use Layer was used to represent zoning with the field Max Dwelling Units per Acre.

So from a technical standpoint, we are looking into:

### Does the Maximum Dwelling Units per Acre in a Block Group correlate with the Dissimilarity Index of the Block Group?

### 

The segregation measure is a set of dissimilarity indices between racial groups. Between any two groups a measure of dissimilarity is calculated for each block group. More documentation can be found here: <https://www.censusscope.org/about_dissimilarity.html>

```{r data_prep, echo=FALSE, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
library(corrr)
library(writexl)
library(openxlsx)
library(ggplot2)
library(tidyverse)
library(sf)
library(leaflet)
library(tidyverse)
library(sf)
library(leaflet)
library(tidycensus)
library(writexl)
library(htmlwidgets)
library(Hmisc)

setwd("~/GitHub/data-science/dissimilarity")


flu<- st_read('C:/Users/SChildress/Documents/ReferenceResearch/displacement/diss_flu16.shp')

flu_df<-as.data.frame(flu)
dissim_index <- read.csv('dissim_bg.csv')

flu_df$GEOID10<- as.numeric(flu_df$GEOID10)
flu_df<-flu_df%>%drop_na(GEOID10)

```

Because the future land use layer and the block groups do not match up geographically, I needed to do an aggregation to get a one to one match between the future land use and the block groups. For this, I found the max, min, and mean maximum du for a block group for all intersecting flu areas.

```{r summarize_vars, echo=FALSE}
flu_df<-flu_df%>%group_by(GEOID10)%>% arrange(desc(max_du_ac))%>%summarise(max_max_du=first(max_du_ac), min_max_du=last(max_du_ac), mean_max_du=mean(max_du_ac))
# remove 0s, these represent geographic mismatches
flu_df <- flu_df %>% filter(max_max_du!=0)
flu_dissim<-merge(flu_df, dissim_index, by ="GEOID10")


describe(flu_dissim$max_max_du)
describe(flu_dissim$min_max_du)
describe(flu_dissim$mean_max_du)

describe(flu_dissim$White_Black_Dissim)
describe(flu_dissim$White_Minority_Dissim)

ggplot(flu_dissim, aes(x=max_max_du)) + geom_histogram(bins=100)
ggplot(flu_dissim, aes(x=min_max_du)) + geom_histogram(bins=100)
ggplot(flu_dissim, aes(x=mean_max_du)) + geom_histogram(bins=100)
ggplot(flu_dissim, aes(x=White_Black_Dissim)) + geom_histogram(bins=100)
ggplot(flu_dissim, aes(x=White_Minority_Dissim)) + geom_histogram(bins=100)


```

There are long tails on the max dus. I'm going to cap them at 100. I think I'll use max_max_du at this point too.

```{r more_analysis}

flu_dissim<- flu_dissim %>%mutate(max_max_du_capped=replace(max_max_du,max_max_du>100,100))

ggplot(flu_dissim, aes(x=max_max_du_capped)) + geom_histogram(bins=100)

ggplot(flu_dissim, aes(x=max_max_du_capped, y=White_Black_Dissim))+
  geom_point()+
  geom_smooth(method=lm, formula='y ~poly(x,3)')

ggplot(flu_dissim, aes(x=max_max_du_capped, y=White_Minority_Dissim))+
  geom_point()+
  geom_smooth(method=lm, formula='y ~poly(x,3)')

```

It looks like:

With max DUs ranging from 0 to 25, as the max DUs increases, the White-Black Dissimilarity decreases, and gets closer to 0, meaning less White shares, and more integration.

Also with max Dus ranging from 75 to 100 as the max DUs increases, the White-Black Dissimilarity decreases, and becomes more negative, meaning less White shares, but more segregation of People of Color (I think this is what it means).

```{r stat tests}

cor.test(flu_dissim$White_Minority_Dissim, flu_dissim$max_max_du,
         method = "spearman")

cor.test(flu_dissim$White_Black_Dissim, flu_dissim$max_max_du,
         method = "spearman")

cor.test(flu_dissim$White_AIAN_Dissim, flu_dissim$max_max_du,
         method = "spearman")

```

On the total range of DUs per acre the correlation is weak.

```{r thresholds}

sf_threshold_low<-6



# Add a variable for defining single family as greater than 0, but less than some max DU per acre
flu_dissim<-flu_dissim %>% mutate(
  is_sf_low = case_when(
    max_max_du>=sf_threshold_low ~ 0,
    max_max_du<sf_threshold_low ~ 1
  ))%>%
  mutate(is_sf_factor_low = as.numeric(is_sf_low))


mean_low<-flu_dissim %>%
  group_by(is_sf_low)%>%
  summarise(mean_white_minority_dissim_low=mean(White_Minority_Dissim))

flu_dissim %>%
  group_by(is_sf_low)%>%
  summarise(mean_white_black_dissim_low=mean(White_Black_Dissim))

flu_dissim %>%
  group_by(is_sf_low)%>%
  summarise(mean_white_aian_dissim_low=mean(White_AIAN_Dissim))


cor.test(flu_dissim$White_Minority_Dissim, flu_dissim$is_sf_low,
         method = "spearman")

cor.test(flu_dissim$White_Black_Dissim, flu_dissim$is_sf_low,
         method = "spearman")
cor.test(flu_dissim$White_AIAN_Dissim, flu_dissim$is_sf_low,
         method = "spearman")


anova_sf_white_black = aov(White_Black_Dissim ~is_sf_low, data=flu_dissim)
anova_sf_white_minority = aov(White_Black_Dissim ~is_sf_low, data=flu_dissim)
anova_sf_white_aian = aov(White_AIAN_Dissim ~is_sf_low, data=flu_dissim)
summary(anova_sf_white_aian)
summary(anova_sf_white_minority)
summary(anova_sf_white_black)

```

This is not producing much correlation. What if we try a multibreak point approach: 0-3 DUs low sf, 3-6 med sf, 6-10 high sf, 10-25 low mf, 25-75 med mf, 75+ high mf.

```{r multibreak approach}

low_sf<-3
med_sf<-6
high_sf<-10
low_mf<-25
med_mf<-75


# Add a variable for defining single family as greater than 0, but less than some max DU per acre
flu_dissim<-flu_dissim %>% mutate(sf_mf_cats = case_when(
    max_max_du<=low_sf ~ "Low_SF",
    max_max_du<=med_sf ~ "Med_SF",
    max_max_du<=high_sf ~ "High_SF",
    max_max_du<=low_mf ~ "Low_MF",
    max_max_du<=med_mf ~ "Med_MF",
    max_max_du>=med_mf ~ "High_MF"
  ))

flu_dissim$sf_mf_cats <-factor(flu_dissim$sf_mf_cats, levels=c("Low_SF", "Med_SF",
                                                               "High_SF", "Low_MF",
                                                               "Med_MF", "High_MF"))

flu_dissim %>% group_by(sf_mf_cats) %>% summarise(mean_dissim_white_black=mean(White_Black_Dissim))

du_cat_dissim<-ggplot(flu_dissim, aes(x=sf_mf_cats, y=White_Black_Dissim))+geom_boxplot()

du_cat_dissim
```

It looks like a few categories can be collapsed. Med_SF/High_SF,

```{r multibreak2}
  
low_sf<-3
med_high_sf<-10
low_mf<-25
med_mf<-75


# Add a variable for defining single family as greater than 0, but less than some max DU per acre
flu_dissim<-flu_dissim %>% mutate(sf_mf_cats = case_when(
    max_max_du<=low_sf ~ "Low_SF (less than 3 Du per acre)",
    max_max_du<=med_high_sf ~ "Med_High_SF (3-10)",
    max_max_du<=low_mf ~  "Low_MF (10-25)",
    max_max_du<=med_mf ~ "Med_MF(25-75)",
    max_max_du>=med_mf ~ "High_MF(75+)"
  ))

flu_dissim$sf_mf_cats <-factor(flu_dissim$sf_mf_cats, levels=c("Low_SF (less than 3 Du per acre)",
                                                               "Med_High_SF (3-10)",
                                                                "Low_MF (10-25)",
                                                               "Med_MF(25-75)",
                                                               "High_MF(75+)"))

flu_dissim %>% group_by(sf_mf_cats) %>% summarise(mean_dissim_white_black=mean(White_Black_Dissim))

du_cat_dissim<-ggplot(flu_dissim, aes(x=sf_mf_cats, y=White_Black_Dissim))+geom_boxplot()


du_cat_dissim_aian<-ggplot(flu_dissim, aes(x=sf_mf_cats, y=White_AIAN_Dissim))+geom_boxplot()
```

```{r estimate_model_with_more_Vars}
acs_rent_inc = read.csv("C:/Users/SChildress/Documents/GitHub/data-science/dissimilarity/census_tract_vars.csv")


# housing_policy<-read.xlsx("C:/Users/SChildress/Documents/GitHub/data-science/HHSurvey/estimation_displace/housing_policy_city.xlsx")
# 
# housing_policy[is.na(housing_policy)]<-0
# housing_policy[housing_policy=='x']<-1


flu_dissim$GEOID10 <- as.character(flu_dissim$GEOID10)
acs_rent_inc$GEOID <- as.character(acs_rent_inc$GEOID)
flu_dissim <- merge(flu_dissim,acs_rent_inc, by.x='GEOID10', by.y='GEOID', all.x=TRUE)



```

```{r}
flu_dissim_model<-glm(White_Black_Dissim ~ sf_mf_cats,
                 data=flu_dissim)


```

<https://datascience.stackexchange.com/questions/893/how-to-get-correlation-between-two-categorical-variable-and-a-categorical-variab>

Looking to the correlation between the categories and the dissimiliarity index with a statistical test.

```{r anova}

anova = aov(White_Black_Dissim ~sf_mf_cats, data=flu_dissim)
summary(anova)
mean_low
```
