---
title: "Telework"
author: "Polina Butrina"
date: "`r format(Sys.time(), '%B %d %Y')`"
output: 
  html_document:
    df_print: paged
    toc: true
    toc_depth: 6
    toc_float: true
  word_document: default
  pdf_document: default
---

This report presents preliminary findings for telework time trends based on the Household Travel Survey (2017 and 2019).

```{r global-options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, 
                      warning=FALSE, 
                      message=FALSE) # formatting
```

```{r Libraries, include=FALSE}
library(nnet)
library(data.table)
library(tidyverse)
library(DT)
library(openxlsx)
library(odbc)
library(DBI)

library(Hmisc) # before dplyr so dplyr functions will worklibrary(Hmisc)
library(corrplot)

library(dplyr)
library(summarytools)
#library(stargazer)
library(MASS)
#library(caret) #requires lattice and ggplot2
library(ggplot2)
library(lattice)
#library(interactions)
#library(hablar)
library(kableExtra)

library(tidyr)
#library(reshape2)
library(table1)
library(knitr)
#library(kableExtra)
```
```{r Functions}
p_MOE <- 0.5
z<-1.645
missing_codes <- c('Missing: Technical Error', 'Missing: Non-response', 
                   'Missing: Skip logic', 'Children or missing', 'Prefer not to answer')

# connecting to Elmer
db.connect <- function() {
  elmer_connection <- dbConnect(odbc(),
                                driver = "SQL Server",
                                server = "AWS-PROD-SQL\\Sockeye",
                                database = "Elmer",
                                trusted_connection = "yes"
  )
}

# a function to read tables and queries from Elmer
read.dt <- function(astring, type =c('table_name', 'sqlquery')) {
  elmer_connection <- db.connect()
  if (type == 'table_name') {
    dtelm <- dbReadTable(elmer_connection, SQL(astring))
  } else {
    dtelm <- dbGetQuery(elmer_connection, SQL(astring))
  }
  dbDisconnect(elmer_connection)
  dtelm
}

create_table_one_var = function(var1, table_temp,table_type ) {
  #table_temp = recategorize_var_upd(var2,table_temp)
  #print(table_temp)
  if (table_type == "household" | table_type == "person" ) {
    weight_2017 = "hh_wt_revised"
    weight_2019 = "hh_wt_2019"
    weight_comb = "hh_wt_combined"
  } else if (table_type == "trip") {
    weight_2017 = "trip_weight_revised"
    weight_2019 = "trip_wt_2019"
    weight_comb = "trip_wt_combined"  
  } 
  
  temp = table_temp %>% dplyr::select(!!sym(var1), all_of(weight_2017), all_of(weight_2019), all_of(weight_comb)) %>% 
    filter(!.[[1]] %in% missing_codes, !is.na(.[[1]])) %>% 
    group_by(!!sym(var1)) %>%  
    summarise(n=n(),sum_wt_comb = sum(.data[[weight_comb]],na.rm = TRUE),sum_wt_2017 = sum(.data[[weight_2017]],na.rm = TRUE),sum_wt_2019 = sum(.data[[weight_2019]],na.rm = TRUE)) %>% 
    mutate(perc_comb = sum_wt_comb/sum(sum_wt_comb)*100, perc_2017 = sum_wt_2017/sum(sum_wt_2017)*100, perc_2019 = sum_wt_2019/sum(sum_wt_2019)*100,delta = perc_2019-perc_2017) %>% 
    ungroup() %>%  mutate(MOE=z*(p_MOE/sum(n))^(1/2)*100) %>% arrange(desc(perc_comb))
  return(temp)
}

create_table_one_var_trip = function(var1, table_temp,table_type ) {
  #table_temp = recategorize_var_upd(var2,table_temp)
  #print(table_temp)
  if (table_type == "household" | table_type == "person" ) {
    weight_2017 = "hh_wt_revised"
    weight_2019 = "hh_wt_2019"
    weight_comb = "hh_wt_combined"
  } else if (table_type == "trip") {
    weight_2017 = "trip_weight_revised"
    weight_2019 = "trip_wt_2019"
    weight_comb = "trip_wt_combined"  
  } else if (table_type == "day") {
    weight_2017 = "hh_day_wt_revised"
    weight_2019 = "hh_day_wt_2019"
    weight_comb = "hh_day_wt_combined"
  }
  
 # trip_wt = "shop_trip"
  
  temp = table_temp %>% dplyr::select(!!sym(var1), all_of(weight_2017), all_of(weight_2019), all_of(weight_comb)) %>% 
    filter(!.[[1]] %in% missing_codes, !is.na(.[[1]])) %>% 
    group_by(!!sym(var1)) %>% 
    summarise(n=n(),sum_wt_comb = sum(.data[[weight_comb]],na.rm = TRUE),sum_wt_2017 = sum(.data[[weight_2017]],na.rm = TRUE),sum_wt_2019 = sum(.data[[weight_2019]],na.rm = TRUE)) %>% 
    mutate(perc_comb = sum_wt_comb/sum(sum_wt_comb)*100, perc_2017 = sum_wt_2017/sum(sum_wt_2017)*100, perc_2019 = sum_wt_2019/sum(sum_wt_2019)*100,delta = perc_2019-perc_2017) %>% 
    ungroup() %>%  mutate(MOE=z*(p_MOE/sum(n))^(1/2)*100) %>% arrange(desc(perc_comb))
  return(temp)
}

create_table = function(var1,var2, table_temp ) {
  #table_temp = recategorize_var_upd(var2,table_temp)
  #print(table_temp)
  temp = table_temp %>% dplyr::select(!!sym(var1), !!sym(var2),hh_wt_revised,hh_wt_2019, hh_wt_combined) %>% 
    filter(!.[[1]] %in% missing_codes, !is.na(.[[1]]),!is.na(.[[2]]),!.[[2]] %in% missing_codes) %>% 
    group_by(!!sym(var1), !!sym(var2)) %>% 
    summarize(n=n(),sum_wt_comb = sum(hh_wt_combined,na.rm = TRUE),sum_wt_2017 = sum(hh_wt_revised,na.rm = TRUE),sum_wt_2019 = sum(hh_wt_2019,na.rm = TRUE)) %>% 
    group_by(!!sym(var2)) %>%  
    mutate(perc_comb = sum_wt_comb/sum(sum_wt_comb)*100, perc_2017 = sum_wt_2017/sum(sum_wt_2017)*100, perc_2019 = sum_wt_2019/sum(sum_wt_2019)*100,delta = perc_2019-perc_2017) %>% 
    ungroup() %>%  mutate(MOE=1.65*(0.25/sum(n))^(1/2)*100)
  return(temp)
  
  
}

#function to parse time categories (characters) to number of minutes (int)
time_parse = function (value) {
  
  if (is.na(value) == TRUE ) {
    return (NA)
  } else if (value == "None"){
    return (0)
  }
  temp = str_split (value, " ")[[1]]
  
  if (length(temp) == 4 ) {
    return ( as.numeric(temp[1]) *60 + as.numeric(temp[3])/60)
  } else if (temp[2] == "hour") {
    return(60)
  } else if (temp[2] == "hours") {
    return( 60 * as.numeric(temp[1]))
  } else if (temp[2] == "minutes") {
    return( as.numeric(temp[1])/60)
  }
  
}

cross_tab_categorical <- function(table, var1, var2, wt_field) {
  expanded <- table %>% 
    group_by(.data[[var1]],.data[[var2]]) %>%
    summarise(Count= n(),Total=sum(.data[[wt_field]])) %>%
    group_by(.data[[var1]])%>%
    mutate(Share=Total/sum(Total)*100)
  
  
  expanded_pivot <-expanded%>%
    pivot_wider(names_from=.data[[var2]], values_from=c(Share,Total, Count))
  
  return (expanded_pivot)
  
} 


# create stargazer table
stargazer_table <- function(outcome, explanatory, table_title){
  stargazer::stargazer(
    polr(as.factor(outcome) ~ explanatory, Hess=T), type = "html",
                      title =  table_title,
                      notes.append =  FALSE, 
                      notes =  c("<sup>&sstarf;</sup>p<0.1; <sup>&sstarf;&sstarf;</sup>p<0.05; <sup>&sstarf;&sstarf;&sstarf;</sup>p<0.01"))

}

# rounding function applied to categorical columns 
my.render.cat <- function(x) {
    c("", sapply(stats.default(x), function(y) with(y,
        sprintf("%d (%.0f%%)", FREQ, PCT))))
  }
```

```{r Data setup, include = FALSE, cache= TRUE}


#read in the household table
elmer_connection <- dbConnect(odbc::odbc(),
                              driver = "SQL Server",
                              server = "AWS-PROD-SQL\\Sockeye",
                              database = "Elmer",
                              trusted_connection = "yes")
h <- dbGetQuery(elmer_connection,
                "SELECT * FROM HHSurvey.v_households_2017_2019_in_house")
p <- dbGetQuery(elmer_connection,
                "SELECT * FROM HHSurvey.v_persons_2017_2019_in_house")

d <- dbGetQuery(elmer_connection,
                "SELECT * FROM HHSurvey.v_days_2017_2019_public")

t <- dbGetQuery(elmer_connection,
                "SELECT * FROM HHSurvey.v_trips_2017_2019_public")

dbDisconnect(elmer_connection)


household <- data.table(h)
person <- data.table(p)
days <- data.table(d)
trips <- data.table(t)

# Statistical assumptions for margins of error
p_MOE <- 0.5
z<-1.645
missing_codes <- c('Missing: Technical Error', 'Missing: Non-response', 
                   'Missing: Skip logic', 'Children or missing', 'Prefer not to answer',
                   'Missing')
```
```{r Data Prep and Data Joins, include = FALSE, cache= TRUE}
#creating a new column 
days_telecom = days %>% mutate(upd_telework_time = 0)

for (i in 1:nrow(days)){
  days_telecom$upd_telework_time[i] = time_parse(days$telework_time[i])
}
person_simple = person %>% dplyr::select(-c(weighted_trip_count_revised,weighted_trip_count_combined,weighted_trip_count_2019, hh_wt_revised,hh_wt_2019,hh_wt_combined,
                                     hh_day_wt_2019, hh_day_wt_combined,hh_day_wt_revised))
household_simple = household %>% dplyr::select(-c( hh_wt_revised,hh_wt_2019,hh_wt_combined,
                                           hh_day_wt_2019, hh_day_wt_combined,hh_day_wt_revised))
days_telecom_person = left_join(days_telecom, person_simple, by = c("personid" = "person_id"))
days_telecom_person_hh = left_join(days_telecom_person,household_simple, by = c("hhid" = "household_id"))

days_telecom_person_hh = days_telecom_person_hh %>% filter(!is.na(upd_telework_time)) %>% 
  mutate(
    telework_cat = case_when(
      (as.numeric(upd_telework_time)) == 0  ~ "None",
      (as.numeric(upd_telework_time) > 0 & as.numeric(upd_telework_time) <= 60)  ~ "Less than 1 hr",
      (as.numeric(upd_telework_time) > 60 & as.numeric(upd_telework_time) < 360) ~ "Part-time",
      as.numeric(upd_telework_time) >= 360 & as.numeric(upd_telework_time) <= 720 ~ "Full-time or greater",
      TRUE ~"Other"
    )) %>% 
  filter(!telework_cat == "Other")

```

## 1. Socio-economic characteristics
This section describes the socio-economic characteristics of people who telework and people who dont telework.
\
*Summary of telewrok time*
```{r summary}
# freq(person_and_household$hhincome_broad)

days_telecom_person_hh$telework_cat <- factor(days_telecom_person_hh$telework_cat,
                                                levels=c("None","Less than 1 hr","Part-time","Full-time or greater","Other"))

dt = create_table_one_var_trip("telework_cat", days_telecom_person_hh,"day" ) %>%
  mutate (Weighted_Sum = round(sum_wt_comb,0), Share = round(perc_comb,0), MOE = round(MOE,2) ) %>% 
  dplyr::select(telework_cat,n, Weighted_Sum,Share,MOE) %>% arrange(telework_cat)

dt %>%
  kbl() %>%
  #kable_material(c("striped", "hover"))
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))

```


*Telewrok by Age*
```{r age}
dt = cross_tab_categorical(days_telecom_person_hh,"telework_cat","age", "hh_day_wt_combined")
number = nrow(days_telecom_person_hh %>% group_by(age) %>% tally())
dt %>% mutate_at(vars(-telework_cat), funs(round(., 1))) %>% 
  dplyr::select(1:(number+1)) %>% 
  kbl() %>%
  #kable_material(c("striped", "hover"))
  kable_paper("hover", full_width = F)
```


*Telewrok by Income*
```{r income}

days_telecom_person_hh$hhincome_broad = factor(days_telecom_person_hh$hhincome_detailed, levels = c("Under $10,000", "$10,000-$24,999", "$25,000-$34,999", "$35,000-$49,999", "$50,000-$74,999","$75,000-$99,999","$100,000-$149,999", "$150,000-$199,999", "$200,000-$249,999", "$250,000 or more") )

dt = cross_tab_categorical(days_telecom_person_hh,"telework_cat","hhincome_detailed", "hh_day_wt_combined") %>% arrange(telework_cat)
#number = nrow(days_telecom_person_hh %>% group_by(hhincome_detailed) %>% tally())
col_order <- c("telework_cat","Share_Under $10,000", "Share_$10,000-$24,999", "Share_$25,000-$34,999", "Share_$35,000-$49,999", "Share_$50,000-$74,999","Share_$75,000-$99,999","Share_$100,000-$149,999", "Share_$150,000-$199,999", "Share_$200,000-$249,999", "Share_$250,000 or more", "Share_Prefer not to answer")
dt <- dt[, col_order]

dt %>% #mutate_at(vars(-telework_cat), funs(round(., 1)))%>% 
  dplyr::select(1:(number+1)) %>% 
  kbl() %>%
  #kable_material(c("striped", "hover"))
  kable_paper("hover", full_width = F)
```
*Telewrok by Gender*
```{r gender}

dt = cross_tab_categorical(days_telecom_person_hh,"telework_cat","gender", "hh_day_wt_combined") %>% arrange(telework_cat)
number = nrow(days_telecom_person_hh %>% group_by(gender) %>% tally())

dt %>% mutate_at(vars(-telework_cat), funs(round(., 1)))%>% 
  dplyr::select(1:(number+1)) %>% 
  kbl() %>%
  #kable_material(c("striped", "hover"))
  kable_paper("hover", full_width = F)

```

```{r education}

dt = cross_tab_categorical(days_telecom_person_hh,"telework_cat","education", "hh_day_wt_combined") %>% arrange(telework_cat)

number = nrow(days_telecom_person_hh %>% group_by(education) %>% tally())

dt %>% mutate_at(vars(-telework_cat), funs(round(., 1)))%>% 
  dplyr::select(1:(number+1)) %>% 
  kbl() %>%
  #kable_material(c("striped", "hover"))
  kable_paper("hover", full_width = F)

```

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
